Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-10-02 astronum_2019_MAESTROeX/astronum-maestroex.tex
added citations, wrote paragraph about the speedup figure

\newcommand{\maestro}{{\sffamily Maestro}}
\newcommand{\maestroex}{{\sffamily MAESTROeX}}
\newcommand{\castro}{{\sffamily Castro}}
\newcommand{\nyx}{{\sffamily Nyx}}
\newcommand{\amrex}{{\sffamily AMReX}}
\newcommand{\amrexastro}{{\sffamily AMReX-Astro}}
\newcommand{\fboxlib}{{\sffamily FBoxLib}}

@@

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible \cite{Boussinesq1901} and anelastic \cite{Ogura1962a,Gough1968,Durran1989} approximations, however the approximation that we use (called the low Mach number approximation \cite{Day2000,Almgren2006a,Nonaka2010} and the generalized pseudo-incompressible approximation \cite{Vasil2013}) allows for background stratification and large density and temperature perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a 1d hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(r, \bm{x}, t)$, such that the full pressure is given by $p(r, \bm{x}, t) = p_0(r, t) + \pi(\bm{x}, t)$. In the low Mach number regime, asymptotic analysis shows that $|\pi|/p_0 = O(\mathrm{Ma}^2)$.

@@

Prior to \maestroex, we developed the low Mach number code \maestro. Like \maestroex, \maestro~is a block-structured adaptive mesh refinement (AMR) code for modelling low Mach number astrophysical codes. Its development is described in a series of papers: \cite{Almgren2006a,Almgren2006b,Almgren2008a,Zingale2009,Nonaka2010}. In \maestro, the system of low Mach number equations is solved using an explicit Godunov approach for the advection, a stiff ODE solver for the reactions (VODE~\cite{vode}), and multigrid linear solvers for the pressure projection steps. Strang splitting CITE is used to integrate the thermodynamic variables, a second order projection method to integrate the velocity subject to the divergence constraint, and a velocity splitting scheme to hydrodynamically evolve the base state. The original \maestro~code was developed using the pure Fortran 90 \fboxlib~software framework \cite{Zhang2016}; \maestroex~instead uses the C++/Fortran 90 \amrex~framework \cite{Zhang2019}.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code \cite{Almgren2010}, and \nyx, an N-body hydrodynamics cosmological simulation code \cite{Almgren2013}. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. A prime example of this is the GPU port capability that has recently been implemented in \castro, and are currently in the process of being implemented in \maestroex. 
\MarginPar{mention starkiller astrophysics}

@@

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learned porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code to required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.   
In order to achieve the best performance on the GPU, we require some additional optimization of the computational kernals. However, we find that these optimizations often improve the performance when the code is run on the CPU only as well. 
More details to appear in a forthcoming \amrexastro~GPU paper.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{plots/gpu_speedup}
    \caption{Comparison of the performance of functions on the CPU vs the GPU. The plot shows the average execution time for several functions, recorded for a run of the 3d \texttt{reacting\_bubble} problem on Summit. Both tests were performed using a single node, with the CPU test using 42 MPI processes, each with 4 OpenMP threads, and the GPU test using 6 MPI processes, each process associated with a single GPU. Note that the $x$-axis scale is logarithmic.}\label{fig:gpu_speedup}
\end{figure}

@@

\ack Figure~\ref{fig:gpu_speedup} was generated using \texttt{Jupyter}
\citep{Kluyver2016} and \texttt{matplotlib} \citep{Hunter2007}.  The
work at Stony Brook was supported by the SciDAC program DOE grant
DE-SC0017955 and DOE/Office of Nuclear Physics grant
DE-FG02-87ER40317.  The work at LBNL was supported by the DOE Office
of Advanced Scientific Computing Research under Contract No,
DE-AC02-05CH11231.  An award of computer time was provided by the
Innovative and Novel Computational Impact on Theory and Experiment
(INCITE) program. This research used resources of the Oak Ridge
Leadership Computing Facility at the Oak Ridge National Laboratory,
which is supported by the Office of Science of the U.S. Department of
Energy under Contract No.\ DE-AC05-00OR22725.  This research used
resources of the National Energy Research Scientific Computing Center,
which is supported by the Office of Science of the U.S. Department of
Energy under Contract No.\ DE-AC02-05CH11231. This research has made
use of NASA's Astrophysics Data System Bibliographic Services.

>>>>

\newcommand{\maestro}{{\sffamily Maestro}}
\newcommand{\maestroex}{{\sffamily MAESTROeX}}
\newcommand{\castro}{{\sffamily Castro}}
\newcommand{\nyx}{{\sffamily Nyx}}
\newcommand{\amrex}{{\sffamily AMReX}}
\newcommand{\starkiller}{{\sffamily Starkiller Microphysics}}
\newcommand{\amrexastro}{{\sffamily AMReX-Astro}}
\newcommand{\fboxlib}{{\sffamily FBoxLib}}

@@

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible \cite{Boussinesq1901} and anelastic \cite{Ogura1962a,Gough1968,Durran1989} approximations, however the approximation that we use (called the low Mach number approximation \cite{Day2000,Almgren2006a,Nonaka2010} and the generalised pseudo-incompressible approximation \cite{Vasil2013}) allows for background stratification and large density and temperature perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a 1d hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(r, \bm{x}, t)$, such that the full pressure is given by $p(r, \bm{x}, t) = p_0(r, t) + \pi(\bm{x}, t)$. In the low Mach number regime, asymptotic analysis shows that $|\pi|/p_0 = O(\mathrm{Ma}^2)$.

@@

Prior to \maestroex, we developed the low Mach number code \maestro. Like \maestroex, \maestro~is a block-structured adaptive mesh refinement (AMR) code for modelling low Mach number astrophysical codes. Its development is described in a series of papers: \cite{Almgren2006a,Almgren2006b,Almgren2008a,Zingale2009,Nonaka2010}. In \maestro, the system of low Mach number equations is solved using an explicit Godunov approach for the advection, a stiff ODE solver for the reactions (VODE, \cite{vode}), and multigrid linear solvers for the pressure projection steps. Strang splitting \cite{strang1968} is used to integrate the thermodynamic variables, a second order projection method to integrate the velocity subject to the divergence constraint, and a velocity splitting scheme to hydrodynamically evolve the base state. The original \maestro~code was developed using the pure Fortran 90 \fboxlib~software framework \cite{Zhang2016}; \maestroex~instead uses the C++/Fortran 90 \amrex~framework \cite{Zhang2019}.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code \cite{Almgren2010}, and \nyx, an N-body hydrodynamics cosmological simulation code \cite{Almgren2013}. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. \castro~and \maestroex~share a common set of microphysics solvers provided by the \starkiller~library\footnote{\url{https://github.com/starkiller-astro/Microphysics}}. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. A prime example of this is the GPU port capability that has recently been implemented in \castro, and are currently in the process of being implemented in \maestroex. 
% \MarginPar{mention starkiller astrophysics}

@@

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learned porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code to required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.   
In order to achieve the best performance on the GPU, we require some additional optimization of the computational kernels. However, we find that these optimizations often improve the performance when the code is run on the CPU only as well. 
More details of the approach used to port our codes to GPU are to appear in a forthcoming \amrexastro~GPU paper.

@@

As mentioned, our work porting \maestroex~to run on GPUs is still ongoing, however in \ref{fig:gpu_speedup}, we show the speedup we have achieved so far for a number of individual functions that have been offloaded to GPU. The plot shows the average execution time of each function, recorded for the 3d \texttt{reacting\_bubble} problem on the OLCF Summit machine. Both tests were run on a single node, consisting of two IBM POWER9 processors (together providing 42 physical CPU cores) and six NVIDIA Volta GPUs. The CPU test used 42 MPI processes, each with 4 OpenMP threads, and the GPU test used 6 MPI processes, each process associated with a single GPU. Note that the plot's $x$-axis scale is logarithmic. As can be seen, large speedups in excess of 30-40$\times$ were achieved for several of the functions when run on the GPU. Once we have finished porting the remaining routines to GPU and further optimised our code, we hope to achieve this sort of speedup for the entire timestep.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{plots/gpu_speedup}
    \caption{Comparison of the performance of functions on the CPU vs the GPU. The plot shows the average execution time for several functions, recorded for a run of the 3d \texttt{reacting\_bubble} problem on Summit. Both tests were performed using a single node, with the CPU test using 42 MPI processes, each with 4 OpenMP threads, and the GPU test using 6 MPI processes, each process associated with a single GPU. Note that the $x$-axis scale is logarithmic. On the right, we have listed the speedup for each function when run on GPUs.}\label{fig:gpu_speedup}
\end{figure}

@@

\ack Figure~\ref{fig:gpu_speedup} was generated using \texttt{Jupyter} \citep{Kluyver2016} and \texttt{matplotlib} \citep{Hunter2007}.
